A nice acyclic matching on the nerve of the partition lattice

### 2019-07-15 paper3/equivariant.tex
extract for all, mistake in paper2
k, n-k

It is well-known that $\Delta(\overline{\Pi}_n)$ is homotopy equivalent to a wedge of spheres of dimension $n-3$. The following two theorems are the first results concerning the topology of the quotient $\Delta(\overline{\Pi}_n)/G$, where $G$ is a non-trivial subgroup of $S_n$.
\begin{thm}[Kozlov, \cite{clmap}] % p. 198
For any $n\geq 3$, the topological space $\Delta(\overline{\Pi}_n)/S_n$ is contractible.
\end{thm}
We consider the Young subgroup $S_1\times S_{n-1}:=\{\sigma\in S_n\mid\sigma(1)=1\}$.
\begin{thm}[Donau, \cite{donau}]
\label{rdthm}
Let $n\geq3$ and $G\subset S_1\times S_{n-1}$ be a subgroup, then the topological space $\Delta(\overline{\Pi}_n)/G$ is homotopy equivalent to a wedge of $k$ spheres of dimension $n-3$, where $k$ is the index of $G$ in $S_1\times S_{n-1}$.
\end{thm}
This leads to a general question of determining the homotopy type of $\Delta(\overline{\Pi}_n)/G$ for an arbitrary subgroup $G\subset S_n$. One might conjecture that $\Delta(\overline{\Pi}_n)/G$ is homotopy equivalent to a wedge of spheres for any $n\geq 3$ and any subgroup $G\subset S_n$. But unfortunately this statement is not true as the following example will show: Let $p\geq 5$ be a prime number and let $C_p$ denote the subgroup of $S_p$ that is generated by the cycle $(1,2,\dots,p)$. Then the fundamental group of $\Delta(\overline{\Pi}_p)/C_p$ is isomorphic to $\Z/p\Z$. In particular $\Delta(\overline{\Pi}_p)/C_p$ cannot be homotopy equivalent to a wedge of spheres. A proof, which uses facts about covering spaces\footnote{See \cite[Chapter 1.3]{hatcher}}, can be found in \cite{torsion}.

@@

Consider the Young subgroup $S_{k_1}\times S_{k_2}\times\dots\times S_{k_r}\subset S_n$ with $k_1+k_2+\dots+k_r=n$. During my work on this paper Gregory Arone\footnote{See Example 1.4 in \cite{arone}} discovered the following: $\Delta(\overline{\Pi}_n)$ is $(S_{k_1}\times S_{k_2}\times\dots\times S_{k_r})$-homotopy equivalent to a wedge of $(n-1)!$ spheres of dimension $n-3$ if $\gcd(k_1,k_2,\dots,k_r)=1$. $S_{k_1}\times S_{k_2}\times\dots\times S_{k_r}$ acts freely on these spheres. This is a generalization of Corollary \ref{moeglichkeit2} and Lemma \ref{freetrans}. Notice that a preprint of this paper was already published on arXiv in 2012.
\section{Discrete Morse Theory}
The definitions of regular trisps, partial matchings, acyclic matchings and foundations of Discrete Morse Theory can be found in \cite{forman,clmap,buch}. The following two theorems of Discrete Morse Theory are frequently used in our proofs.
\begin{thm}[Patchwork Theorem]
\label{patchwork}
Let $\varphi:P\longrightarrow Q$ be an order-preserving map and assume we have acyclic matchings on the subposets $\varphi^{-1}(q)$ for all $q\in Q$. Then the union of these matchings is an acyclic matching on $P$.
\end{thm}
\begin{thm}%[Main Theorem of Discrete Morse Theory for CW Complexes]
\label{morsemain}
Let $\Delta$ be a finite regular trisp and let $M$ be an acyclic matching on the poset ${\cal F}(\Delta)\setminus\{\hat{0}\}$. Let $c_i$ denote the number of critical $i$-dimensional simplices of $\Delta$. Then $\Delta$ is homotopy equivalent to a CW complex with $c_i$ cells of dimension $i$.
\end{thm}
The proofs of Theorems \ref{patchwork} and \ref{morsemain} as well as further facts on Discrete Morse Theory can be found in \cite[Chapter 11]{buch}.
\section{An Equivariant Patchwork Theorem}
We wish to construct an equivariant acyclic matching on a poset by gluing together smaller equivariant acyclic matchings on parts of the poset. This is similar to Theorem \ref{patchwork} with the difference that we also create copies of these matchings in our construction, see Figure \ref{eqex}.
\begin{df}
Let $P$ be a poset and let $G$ be a group acting on $P$. Let $M$ be an acyclic matching on $P$. We call $M$ an \emph{$G$-equivariant acyclic matching} if $(a,b)\in M$ implies $(ga,gb)\in M$ for all $g\in G$ and $a,b\in P$.
\end{df}
Let $G$ be a group acting on some posets $P$ and $Q$. For an element $q\in Q$ we set $G_q:=\{g\in G\mid gq=q\}$, known as the stabilizer subgroup of $q$.
\begin{prop}
\label{eqpatchwork}
Let $\varphi:P\longrightarrow Q$ be an order-preserving $G$-map and let $R\subset Q$ be a subset such that $R$ contains exactly one representative for each orbit in $Q$. Assume for each $r\in R$ we have an $G_r$-equivariant acyclic matching $M_r$ on $\varphi^{-1}(r)$. For $r\in R$, let $C_r$ denote the set of critical elements of $M_r$. Then we have an $G$-equivariant acyclic matching on $P$ such that
\[
\bigcup_{g\in G, r\in R}gC_r
\]
is the set of critical elements.

@@

For an order-preserving map with small fibers $\varphi$ let $M(\varphi)$ denote its associated acyclic matching which consists of all fibers of cardinality $2$, see \cite[Chapter 11]{buch}.
\begin{prop}
Let $G$ be a group acting on a finite poset $P$. For any order-preserving $G$-map $\varphi:P\longrightarrow Q$ with small fibers, the acyclic matching $M(\varphi)$ is $G$-equivariant. On the other hand, any $G$-equivariant acyclic matching $M$ on $P$ can be represented as $M=M(\varphi)$, where $\varphi:P\longrightarrow Q$ is an order-preserving $G$-map with small fibers.
\end{prop}
For two disjoint posets $(P_1,\leq_{P_1})$, $(P_2,\leq_{P_2})$ let $P_1\oplus P_2$ denote the \emph{ordinal sum}\footnote{Called \emph{stack} in \cite{buch}} of $P_1$ and $P_2$, where the set of $P_1\oplus P_2$ is $P_1\cup P_2$ and for two elements $x,y\in P_1\cup P_2$ we set $x\leq y$ if one of the following three conditions is satisfied:
\begin{enumerate}
\item $x,y\in P_1$ and $x\leq_{P_1} y$
\item $x,y\in P_2$ and $x\leq_{P_2} y$
\item $x\in P_1$ and $y\in P_2$
\end{enumerate}
Furthermore if a group $G$ acts on $P_1$ and on $P_2$, then we obtain a canonical action of $G$ on $P_1\oplus P_2$.
\begin{proof}
The first statement is easy to see. Assume we are given an $G$-equivariant acyclic matching $M$ on $P$.

@@

By Remark \ref{card} the cardinality of $C_n$ is $(n-1)!$ which equals the cardinality of $S_1\times S_{n-1}$. Hence the action is free.
\end{proof}
Let $G\subset S_1\times S_{n-1}$ be an arbitrary subgroup.
\begin{rem}
The cardinality of $C_n/G$ is the index of $G$ in $S_1\times S_{n-1}$.
\end{rem}
\begin{proof}
Apply Lemma \ref{freetrans}.
\end{proof}
Now Theorem \ref{rdthm} follows as a corollary. We can either apply Corollary \ref{moeglichkeit1} or Corollary \ref{moeglichkeit2}.
\begin{cor}
The topological space $\Delta(\overline{\Pi}_n)/G$ is homotopy equivalent to a wedge of spheres of dimension $n-3$. The number of spheres is the index of $G$ in $S_1\times S_{n-1}$.
\end{cor}
\section*{Acknowledgments}
The author would like to thank Dmitry N. Kozlov for this interesting problem, Ragnar Freij and Giacomo d'Antonio for the helpful discussions.
\bibliographystyle{model1-num-names}
\begin{thebibliography}{00}
\bibitem{arone}
G. Arone, \textit{A branching rule for partition complexes}, \texttt{arXiv:1508.05086v1 [math.AT]} (2015).
\bibitem{donau2}
R. Donau, \textit{On a quotient topology of the partition lattice with forbidden block sizes}, Topology and its Applications \textbf{159} (8) (2012), pp. 2052-2057.
\bibitem{donau}
R. Donau, \textit{Quotients of the order complex $\Delta(\overline{\Pi}_n)$ by subgroups of the Young subgroup $S_1\times S_{n-1}$}, Topology and its Applications \textbf{157} (16) (2010), pp. 2476-2479.
\bibitem{torsion}
R. Donau, \textit{Quotients of the topology of the partition lattice which are not homotopy equivalent to wedges of spheres}, \texttt{arXiv:1202.4368 [math.AT]} (2012).
\bibitem{forman}
R. Forman, \textit{Morse theory for cell complexes}, Adv. Math. \textbf{134} (1) (1998), pp. 90-145.
\bibitem{freij}
R. Freij, \textit{Equivariant discrete Morse theory}, Discrete Mathematics \textbf{309} (12) (2009), pp. 3821-3829.
\bibitem{hatcher}
A. Hatcher, \textit{Algebraic Topology}, Cambridge University Press, 2008.
\bibitem{clmap}
D.N. Kozlov, \textit{Closure maps on regular trisps}, Topology and its Applications \textbf{156} (15) (2009), pp. 2491-2495.
\bibitem{buch}
D. Kozlov, \textit{Combinatorial Algebraic Topology}, Algorithms and Computation in Mathematics \textbf{21}, Springer-Verlag Berlin Heidelberg, 2008.
\bibitem{wachs}
M.L. Wachs, \textit{On the (co)homology of the partition lattice and the free Lie algebra}, Discrete Mathematics \textbf{193} (1-3) (1998), pp. 287-319.
\end{thebibliography}
\end{document}

>>>>

It is well-known that $\Delta(\overline{\Pi}_n)$ is homotopy equivalent to a wedge of spheres of dimension $n-3$. The following two theorems are results concerning the topology of the quotient $\Delta(\overline{\Pi}_n)/G$, where $G$ is a non-trivial subgroup of $S_n$.
\begin{thm}[Kozlov, \cite{clmap}] % p. 198
For any $n\geq 3$, the topological space $\Delta(\overline{\Pi}_n)/S_n$ is contractible.
\end{thm}
We consider the Young subgroup $S_1\times S_{n-1}:=\{\sigma\in S_n\mid\sigma(1)=1\}$.
\begin{thm}[Donau, \cite{donau}]
\label{rdthm}
Let $n\geq3$ and $G\subset S_1\times S_{n-1}$ be a subgroup, then the topological space $\Delta(\overline{\Pi}_n)/G$ is homotopy equivalent to a wedge of $k$ spheres of dimension $n-3$, where $k$ is the index of $G$ in $S_1\times S_{n-1}$.
\end{thm}
This leads to a general question of determining the homotopy type of $\Delta(\overline{\Pi}_n)/G$ for an arbitrary subgroup $G\subset S_n$. One might conjecture that $\Delta(\overline{\Pi}_n)/G$ is homotopy equivalent to a wedge of spheres for any $n\geq 3$ and any subgroup $G\subset S_n$. But unfortunately this statement is not true as the following example will show: Let $p\geq 5$ be a prime number and let $C_p$ denote the subgroup of $S_p$ that is generated by the cycle $(1,2,\dots,p)$. Then the fundamental group of $\Delta(\overline{\Pi}_p)/C_p$ is isomorphic to $\Z/p\Z$. In particular $\Delta(\overline{\Pi}_p)/C_p$ cannot be homotopy equivalent to a wedge of spheres. A proof, which uses facts about covering spaces\footnote{See \cite[Chapter 1.3]{hatcher}}, can be found in \cite{torsion}.

@@

Consider the Young subgroup $S_k\times S_{n-k}\subset S_n$ with $1\leq k<n$. During my work on this paper Gregory Arone\footnote{See Example 1.4 in \cite{arone}} discovered the following: $\Delta(\overline{\Pi}_n)$ is $(S_k\times S_{n-k})$-homotopy equivalent to a wedge of $(n-1)!$ spheres of dimension $n-3$ if $\gcd(k,n-k)=1$. $S_k\times S_{n-k}$ acts freely on these spheres. This is a generalization of Corollary \ref{moeglichkeit2} and Lemma \ref{freetrans}. Notice that the first preprint of this paper was already published on arXiv in 2012.
\section{Discrete Morse Theory}
The definitions of regular trisps, partial matchings, acyclic matchings and foundations of Discrete Morse Theory can be found in \cite{forman,clmap,buch}. The following two theorems of Discrete Morse Theory are frequently used in our proofs.
\begin{thm}[Patchwork Theorem]
\label{patchwork}
Let $\varphi:P\longrightarrow Q$ be an order-preserving map and assume we have acyclic matchings on the subposets $\varphi^{-1}(q)$ for all $q\in Q$. Then the union of these matchings is an acyclic matching on $P$.
\end{thm}
\begin{thm}%[Main Theorem of Discrete Morse Theory for CW Complexes]
\label{morsemain}
Let $\Delta$ be a finite regular trisp and let $M$ be an acyclic matching on the poset ${\cal F}(\Delta)\setminus\{\hat{0}\}$. Let $c_i$ denote the number of critical $i$-dimensional simplices of $\Delta$. Then $\Delta$ is homotopy equivalent to a CW complex with $c_i$ cells of dimension $i$.
\end{thm}
The proofs of Theorems \ref{patchwork} and \ref{morsemain} as well as further facts on Discrete Morse Theory can be found in \cite[Chapter 11]{buch}.
\section{An Equivariant Patchwork Theorem}
We wish to construct an equivariant acyclic matching on a poset by gluing together smaller equivariant acyclic matchings on parts of the poset. This is similar to Theorem \ref{patchwork} with the difference that we also create copies of these matchings in our construction, see Figure \ref{eqex}.
\begin{df}
Let $P$ be a poset and let $G$ be a group acting on $P$. Let $M$ be an acyclic matching on $P$. We call $M$ an \emph{$G$-equivariant acyclic matching} if $(a,b)\in M$ implies $(ga,gb)\in M$ for all $g\in G$ and $a,b\in P$.
\end{df}
Let $G$ be a group acting on some posets $P$ and $Q$. For an element $q\in Q$ we set $G_q:=\{g\in G\mid gq=q\}$, known as the stabilizer subgroup of $q$.
\begin{prop}
\label{eqpatchwork}
Let $\varphi:P\longrightarrow Q$ be an order-preserving $G$-map and let $R\subset Q$ be a subset such that $R$ contains exactly one representative for each orbit in $Q$. Assume for each $r\in R$ we have an $G_r$-equivariant acyclic matching $M_r$ on $\varphi^{-1}(r)$. For $r\in R$, let $C_r$ denote the set of critical elements of $M_r$. Then we have an $G$-equivariant acyclic matching on $P$ such that
\[
\bigcup_{g\in G, r\in R}gC_r
\]
is the set of critical elements.

@@

For an order-preserving map with small fibers $\varphi$ let $M(\varphi)$ denote its associated acyclic matching which consists of all fibers of cardinality $2$, see \cite[Chapter 11]{buch}.
\begin{prop}
\label{smallfibers}
Let $G$ be a group acting on a finite poset $P$. For any order-preserving $G$-map $\varphi:P\longrightarrow Q$ with small fibers, the acyclic matching $M(\varphi)$ is $G$-equivariant. On the other hand, any $G$-equivariant acyclic matching $M$ on $P$ can be represented as $M=M(\varphi)$, where $\varphi:P\longrightarrow Q$ is an order-preserving $G$-map with small fibers.
\end{prop}
For two disjoint posets $(P_1,\leq_{P_1})$, $(P_2,\leq_{P_2})$ let $P_1\oplus P_2$ denote the \emph{ordinal sum}\footnote{Called \emph{stack} in \cite{buch}} of $P_1$ and $P_2$, where the set of $P_1\oplus P_2$ is $P_1\cup P_2$ and for two elements $x,y\in P_1\cup P_2$ we set $x\leq y$ if one of the following three conditions is satisfied:
\begin{enumerate}
\item $x,y\in P_1$ and $x\leq_{P_1} y$
\item $x,y\in P_2$ and $x\leq_{P_2} y$
\item $x\in P_1$ and $y\in P_2$
\end{enumerate}
Furthermore if a group $G$ acts on $P_1$ and on $P_2$, then we obtain a canonical action of $G$ on $P_1\oplus P_2$.
\begin{proof}
The first statement is easy to see. Assume we are given an $G$-equivariant acyclic matching $M$ on $P$.

@@

By Remark \ref{card} the cardinality of $C_n$ is $(n-1)!$ which equals the cardinality of $S_1\times S_{n-1}$. Hence the action is free.
\end{proof}
Let $G\subset S_1\times S_{n-1}$ be an arbitrary subgroup.
\begin{rem}
The cardinality of $C_n/G$ is the index of $G$ in $S_1\times S_{n-1}$.
\end{rem}
\begin{proof}
Apply Lemma \ref{freetrans}.
\end{proof}
Now Theorem \ref{rdthm} follows as a corollary. We can either apply Corollary \ref{moeglichkeit1} or Corollary \ref{moeglichkeit2}.
\begin{cor}
The topological space $\Delta(\overline{\Pi}_n)/G$ is homotopy equivalent to a wedge of spheres of dimension $n-3$. The number of spheres is the index of $G$ in $S_1\times S_{n-1}$.
\end{cor}
\section*{Acknowledgments}
The author would like to thank Dmitry N. Kozlov for this interesting problem, Ragnar Freij and Giacomo d'Antonio for the helpful discussions.
\bibliographystyle{model1-num-names}
\begin{thebibliography}{00}
\bibitem{arone}
G. Arone, \textit{A branching rule for partition complexes}, \texttt{arXiv:1508.05086 [math.AT]} (2015).
\bibitem{donau2}
R. Donau, \textit{On a quotient topology of the partition lattice with forbidden block sizes}, Topology and its Applications \textbf{159} (8) (2012), pp. 2052-2057.
\bibitem{donau}
R. Donau, \textit{Quotients of the order complex $\Delta(\overline{\Pi}_n)$ by subgroups of the Young subgroup $S_1\times S_{n-1}$}, Topology and its Applications \textbf{157} (16) (2010), pp. 2476-2479.
\bibitem{torsion}
R. Donau, \textit{Quotients of the topology of the partition lattice which are not homotopy equivalent to wedges of spheres}, \texttt{arXiv:1202.4368 [math.AT]} (2012).
\bibitem{forman}
R. Forman, \textit{Morse theory for cell complexes}, Adv. Math. \textbf{134} (1) (1998), pp. 90-145.
\bibitem{freij}
R. Freij, \textit{Equivariant discrete Morse theory}, Discrete Mathematics \textbf{309} (12) (2009), pp. 3821-3829.
\bibitem{hatcher}
A. Hatcher, \textit{Algebraic Topology}, Cambridge University Press, 2008.
\bibitem{clmap}
D.N. Kozlov, \textit{Closure maps on regular trisps}, Topology and its Applications \textbf{156} (15) (2009), pp. 2491-2495.
\bibitem{buch}
D. Kozlov, \textit{Combinatorial Algebraic Topology}, Algorithms and Computation in Mathematics \textbf{21}, Springer-Verlag Berlin Heidelberg, 2008.
\bibitem{wachs}
M.L. Wachs, \textit{On the (co)homology of the partition lattice and the free Lie algebra}, Discrete Mathematics \textbf{193} (1-3) (1998), pp. 287-319.
\end{thebibliography}
\end{document}
